---
title: "Robust backfitting"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A robust backfitting algorithm

An `R` package implementing the robust back-fitting algorithm as proposed in 
[this submitted manuscript](http://www.stat.ubc.ca/~matias/RBF.pdf). 

The package can be installed from within `R` by using
```R
library(devtools)
install_github("msalibian/RBF")
```

Here is a (longish) example on how to use the code. 
We use the Air Quality data.
```{r intro}
library(RBF)
data(airquality)
x <- airquality
x <- x[ complete.cases(x), ]
x <- x[, c('Ozone', 'Solar.R', 'Wind', 'Temp')]
y <- as.vector(x$Ozone)
x <- as.matrix(x[, c('Solar.R', 'Wind', 'Temp')])
```
A scatter plot of the data
```{r scatter, fig.width=7, fig.height=7}
pairs(cbind(y,x), labels=c('Ozone', colnames(x)), pch=19, col='gray30', cex=1.5)
```

The following bandwidths were obtained via a robust 
leave-one-out cross-validation procedure (described in the paper).
Here we just set them to their optimal values:
```{r bandw}
bandw <- c(136.728453,   8.894283,   4.764985)
```
Now we use the robust backfitting algorithm to fit an additive
model using Tukey's bisquare loss (the default tuning
constant for this loss function is 4.685)
```{r rbfone}
fit.full <- backf.rob(Xp=x, yp=y, windows=bandw, epsilon=1e-6, 
                     degree=1, type='Tukey')
```

We display the 3 fits (one per additive component), being
careful with the axis limits
```{r showfits, fig.width=7, fig.height=7}
lim.cl <- lim.rob <- matrix(0, 2, 3)
par(mfrow=c(2,2))
for(j in 1:3) {
  re <- y - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  lim.rob[,j] <- c(min(re), max(re))
  plot(re ~ x[,j], type='p', pch=19, col='gray30', xlab=colnames(x)[j], ylab='', cex=1.5)
  oo <- order(x[,j])
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

We now compute and display the 
classical backfitting fits, with 
bandwidths chosen via leave-one-out CV
```{r classicfits, fig.width=7, fig.height=7}
bandw.cl <- c(91.15, 10.67, 9.53)
fit.cl <- backf.cl(Xp=x, yp=y, windows=bandw.cl, epsilon=1e-6,
                   degree=1)
par(mfrow=c(2,2))
for(j in 1:3) {
  re <- y - fit.cl$alpha - rowSums(fit.cl$g.matrix[,-j])
  lim.cl[,j] <- c(min(re), max(re))
  plot(re ~ x[,j], type='p', pch=19, col='gray30', xlab=colnames(x)[j], ylab='', cex=1.5)
  oo <- order(x[,j])
  lines(x[oo,j], fit.cl$g.matrix[oo,j], lwd=5, col='magenta')
}
```

The following plots are partial residual plots with
both the classical and robust fits 
on them
```{r overlay, fig.width=7, fig.height=7}
lims <- lim.cl
for(j in 1:3) {
  lims[1,j] <- min(lim.cl[1,j], lim.rob[1,j])
  lims[2,j] <- max(lim.cl[2,j], lim.rob[2,j])
}
par(mfrow=c(2,2))
for(j in 1:3) {
  re <- y - fit.cl$alpha - rowSums(fit.cl$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=19, col='gray30', xlab=colnames(x)[j], ylab='', cex=1.5,
      ylim=lims[,j])
  oo <- order(x[,j])
  lines(x[oo,j], fit.cl$g.matrix[oo,j], lwd=5, col='magenta')
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

We look at the residuals from the robust fit to
identify potential outiers
```{r outliers, fig.width=5, fig.height=5}
re.ro <- y - fit.full$alpha - rowSums(fit.full$g.matrix)
ou.ro <- boxplot(re.ro, plot=FALSE)$out
n <- length(re.ro)
ou.ro <- (1:n)[ re.ro %in% ou.ro ]
boxplot(re.ro, col='gray80', pch=19, cex=1.5)
points(rep(1, length(ou.ro)), re.ro[ou.ro], pch=19, cex=2, col='red')
```

We highlight these suspicious observations on the
scatter plot 
```{r showouts, fig.width=7, fig.height=7}
cs <- rep('gray30', nrow(x))
cs[ou.ro] <- 'red'
os <- 1:nrow(x)
os2 <- c(os[-ou.ro], os[ou.ro])
pairs(cbind(y,x)[os2,], labels=c('Ozone', colnames(x)), pch=19, col=cs[os2], cex=1.5)
```

and on the partial residuals plots
```{r showouts2, fig.width=7, fig.height=7}
par(mfrow=c(2,2))
for(j in 1:3) {
  re <- y - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=19, col='gray30', xlab=colnames(x)[j], ylab='', cex=1.5,
       ylim=lims[,j])
  points(re[ou.ro] ~ x[ou.ro,j], pch=19, col='red', cex=1.5)
  oo <- order(x[,j])
  lines(x[oo,j], fit.cl$g.matrix[oo,j], lwd=5, col='magenta')
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

If we use the classical backfitting algorithm on
the data without the potential outliers, we 
obtain almost identical results
```{r bothonclean, fig.width=7, fig.height=7}
# Run the classical backfitting algorithm without outliers
x2 <- x[-ou.ro,]
y2 <- y[-ou.ro]
bandw.cl2 <- c(138.87, 10.52, 4.85)
fit.cl2 <- backf.cl(Xp=x2, yp=y2, windows=bandw.cl2, epsilon=1e-6,
                   degree=1)
par(mfrow=c(2,2))
for(j in 1:3) {
  re <- y - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=19, col='gray30', xlab=colnames(x)[j], ylab='', cex=1.5,
       ylim=lims[,j])
  points(re[ou.ro] ~ x[ou.ro,j], pch=19, col='red', cex=1.5)
  oo <- order(x2[,j])
  lines(x2[oo,j], fit.cl2$g.matrix[oo,j], lwd=5, col='magenta')
  oo <- order(x[,j])
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

