---
title: "Examples"
author: "Martínez and Salibian"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## About this vignette

In this vignette, we will show with a toy example and a real data example how a robust backfitting procedure can be easily implemented.

The `R` package `RBF` (available on CRAN [here](https://cran.r-project.org/package=RBF)) 
implements the robust back-fitting algorithm as proposed by
Boente, Martinez and Salibian-Barrera  in 

> Boente G, Martinez A, Salibian-Barrera M. (2017) Robust estimators for 
> additive models using backfitting. Journal of Nonparametric Statistics. Taylor 
> & Francis; 29, 744-767.
> [DOI: 10.1080/10485252.2017.1369077](https://doi.org/10.1080/10485252.2017.1369077)

This repository contains a development version of `RBF`
which may differ slightly from the one available on CRAN
(until the CRAN version is updated appropriately). 

The package in this repository can be installed from within `R` by using the following code (assuming the [devtools](https://cran.r-project.org/package=devtools)) package is available:
```R
devtools::install_github("msalibian/RBF")
```

Now that the `R` package is downloaded, we can now start to see how this procedure works.

## Let's play with a toy example

The following example corresponds to a 2-dimensional simulated samples with 5% of contaminated responses.

Let begin by defining the additive functions and then generating the simulated sample.

```{r generating data}
function.g1 <- function(x1) 24*(x1-1/2)^2-2
function.g2 <- function(x2) 2*pi*sin(pi*x2)-4
set.seed(140)
n <- 500
x1 <- runif(n)
x2 <- runif(n)
X <- cbind(x1, x2)
eps <- rnorm(n,0,sd=0.15)
regression <- function.g1(x1)+function.g2(x2)
y <- regression + eps
```

The following bandwidths were obtained via a robust K-fold cross-validation procedure. In order to not to lose the focus of what it really matters, that is, the robust estimation of the additive components, here we just set them to their optimal values but the code to obtain this values is a separate section.

```{r bandw1}
bandw <- c(0.05, 0.075)
```

Now we will use the robust backfitting procedure to fit an additive model using the Tukey loss function (with default tuning constant c=1.345) and a linear fit (degree=1) for the estimation procedure at each additive component. In addition, a specific point will be predicted.

```{r robust fit}
library(RBF)
point <- c(0.7, 0.6)
robust.fit <- backf.rob(y ~ X, point=point, windows=bandw, type = 'Tukey', degree=1)
```

The prediction and true values of the additive functions are:

```{r prediction}
robust.fit$prediction
c(function.g1(point[1]), function.g2(point[2]))
```

The following figures plot the partial residuals, the estimated curve (in blue) and the true function (in black) for each additive function:

```{r plots-simu, fig.show="hold", out.width="33%"}
lim.rob <- matrix(0, 2, 2)
functions.g <- cbind(function.g1(X[,1]), function.g2(X[,2]))
for(j in 1:2) {
  res <- y - robust.fit$alpha - robust.fit$g.matrix[,-j]
  lim.rob[,j] <- range(res)
  plot(X[,j], res, type='p', pch=19, col='gray45', xlab=colnames(X)[j], ylab='', 
       cex=1, ylim=lim.rob[,j])
  ord <- order(X[,j])
  lines(X[ord,j], robust.fit$g.matrix[ord,j], lwd=3, col='blue')
  lines(X[ord,j], functions.g[ord,j], lwd=3)
}
```



### With contaminated responses.

In order to appreciate the differences of the estimation procedures when using a classical or robust procedure to estimate the additive components in an additive model, we will now contaminate a 10% of the responses and recalculate both estimators.

```{r contaminated responses}
eps <- rnorm(n,0,sd=0.15)
prop.cont <- 0.10
ou <- rbinom(n, size=1, prob=prop.cont) 
eps.ou <- eps
eps.ou[ ou == 1 ] <- rnorm(sum(ou),mean=15, sd=0.1)
yout <- regression + eps.ou
```

And recompute the bandwidths with the K-fold cross-validation procedure (see the following section for the selected bandwidths). The optimal values are:
```{r bandw-out}
bandw <- c(0.05, 0.075)
```

And obtain the following fit
```{r robust.fit.out}
robust.fit.out <- backf.rob(yout ~ X, point=point, windows=bandw, type = 'Tukey', degree=1)
```
and the following predictions at the fixed point
```{r prediction2}
robust.fit.out$prediction
c(function.g1(point[1]), function.g2(point[2]))
```

Then, we plot the partial residuals, the estimated curves (in blue) and the true functions (in black) for each additive function.

```{r plots-simu2, fig.show="hold", out.width="33%"}
lim.rob <- matrix(0, 2, 2)
functions.g <- cbind(function.g1(X[,1]), function.g2(X[,2]))
for(j in 1:2) {
  res <- yout - robust.fit.out$alpha - robust.fit.out$g.matrix[,-j]
  lim.rob[,j] <- range(res)
  plot(X[,j], res, type='p', pch=19, col='gray45', xlab=colnames(X)[j], ylab='', 
       cex=1, ylim=lim.rob[,j])
  ord <- order(X[,j])
  lines(X[ord,j], robust.fit.out$g.matrix[ord,j], lwd=3, col='blue')
  lines(X[ord,j], functions.g[ord,j], lwd=3)
}
```


It can be appreciated that the estimates state insensitive to the presence of atypical observations in the data.



### Optimal bandwidths

As it was mentioned before, we left to this section all computations related to bandwidth selection for the simulated example.

Since for both clean and contaminated responses we compute robust bandwidths, we will first define a function to compute a K-fold cross-validation procedure. 

As a robust prediction error measure we use `mu^2 + sigma^2` where `mu` and `sigma` are M-estimators of location and scale of
the prediction errors, respectively. The code is copied below:

```{r rkfoldfunction}
backf.rob.cv <- function(k=5, Xp, yp, windows, degree, type) {
  n <- length(yp)
  k1 <- floor(n/k)
  ids <- rep(1:k, each=k1)
  if( length(ids) < n ) ids <- c(ids, 1:(n%%k))
  ids <- sample(ids)
  preds <- rep(NA, n)
  for(j in 1:k) {
    XX <- Xp[ids!=j,]
    yy <- yp[ids!=j]
    tmp <- backf.rob(yy ~ XX, point=Xp[ids==j,], windows=windows, degree=degree, type=type)
    preds[ids==j] <- rowSums(tmp$prediction) + tmp$alpha
  }
  preds.res <- preds - yp
  tmp.re <- RobStatTM::locScaleM(preds.res, na.rm=TRUE)
  sal <- tmp.re$mu^2 + tmp.re$disper^2
  return(sal)
}
```

And so the optimal bandwidths are search over a grid. For the clean instance:

```{r robust.5fold1, eval=FALSE}
# Bandwidth selection with leave-one-out cross-validation
# This takes some time to compute (approx 2 minutes running
# R 3.6.3 on an Intel(R) Core(TM) i7-10900 CPU @ 2.90GHz)
h1 <- c(0.05, 0.075, 0.1, 0.125)
hh <- expand.grid(h1, h1)
nh <- nrow(hh)
rmspe <- rep(NA, nh)
system.time({
for(i in 1:nh){
  print(i)
  rmspe[i] <- backf.rob.cv(k=5, Xp=X, yp=y, windows=hh[i,], degree=1, type='Tukey') 
}
})
i0 <- which.min(rmspe)
bandw <- hh[i0,]
```
```{r optbandw1}
bandw
```
and for the contaminated setting:

```{r robust.5fold2, eval=FALSE}
# Bandwidth selection with leave-one-out cross-validation
# This takes some time to compute (approx 2 minutes running
# R 3.6.3 on an Intel(R) Core(TM) i7-10900 CPU @ 2.90GHz)
h1 <- c(0.05, 0.075, 0.1, 0.125)
hh <- expand.grid(h1, h1)
nh <- nrow(hh)
rmspe <- rep(NA, nh)
system.time({
for(i in 1:nh){
  print(i)
  rmspe[i] <- backf.rob.cv(k=5, Xp=X, yp=yout, windows=hh[i,], degree=1, type='Tukey') 
}
})
i0 <- which.min(rmspe)
bandw <- hh[i0,]
```
```{r optbandw2}
bandw
```


## And now with a real data set

Here is a example on how `RBF` works with a real data set. 
We use the Air Quality data. The interest is in
predicting `Ozone` in terms of three explanatory 
variables: `Solar.R`, `Wind` and `Temp`:
```{r intro}
library(RBF)
data(airquality)
pairs(airquality[, c('Ozone', 'Solar.R', 'Wind', 'Temp')], 
      pch=19, col='gray30', cex=1.5)
```

<!---
The following bandwidths were obtained via a robust 
leave-one-out cross-validation procedure (described in the paper).
As a robust prediction error measure we use `mu^2 + sigma^2` where
`mu` and `sigma` are M-estimators of location and scale of
the prediction errors, respectively. The code can be found in the
README file in the GitHub repository.

The code is copied below:
```{r robust.leaveoneout, eval=FALSE}
# Bandwidth selection with leave-one-out cross-validation
## Without outliers
# This takes a long time to compute (approx 380 minutes running
# R 3.6.1 on an Intel(R) Core(TM) i7-4790 CPU @ 3.60GHz)
ccs <- complete.cases(airquality)
x <- as.matrix( airquality[ccs, c('Solar.R', 'Wind', 'Temp')] )
y <- as.vector( airquality[ccs, 'Ozone'] )
a <- c(1/2, 1, 1.5, 2, 2.5, 3)
h1 <- a * sd(x[,1])
h2 <- a * sd(x[,2])
h3 <- a * sd(x[,3])
hh <- expand.grid(h1, h2, h3)
nh <- nrow(hh)
rmspe <- rep(NA, nh)
jbest <- 0
cvbest <- +Inf
# leave-one-out
n <- nrow(x)
for(i in 1:nh) {
  # leave-one-out CV loop
  preds <- rep(NA, n)
  for(j in 1:n) {
    tmp <- try( backf.rob(y ~ x, point = x[j, ],
                          windows = hh[i, ], epsilon = 1e-6,
                          degree = 1, type = 'Tukey', subset = c(-j) ))
    if (class(tmp)[1] != "try-error") {
      preds[j] <- rowSums(tmp$prediction) + tmp$alpha
    }
  }
  pred.res <- preds - y
  tmp.re <- RobStatTM::locScaleM(pred.res, na.rm=TRUE)
  rmspe[i] <- tmp.re$mu^2 + tmp.re$disper^2
  if( rmspe[i] < cvbest ) {
    jbest <- i
    cvbest <- rmspe[i]
    print('Record')
  }
  print(c(i, rmspe[i]))
}
bandw <- hh[jbest,]
```
--->

Here we just set them to their optimal values:
```{r bandw}
bandw <- c(136.7285, 10.67314, 4.764985)
```
Now we use the robust backfitting algorithm to fit an additive
model using Tukey's bisquare loss (the default tuning
constant for this loss function is 4.685). We remove 
cases with missing entries. 
```{r rbfone}
ccs <- complete.cases(airquality)
fit.full <- backf.rob(Ozone ~ Solar.R + Wind + Temp, data=airquality,
                subset=ccs, windows=bandw, degree=1, type='Tukey')
```

We display the 3 fits (one per additive component), being
careful with the axis limits (which are stored to use them later):
```{r showfits, fig.show="hold", out.width="33%"}
lim.cl <- lim.rob <- matrix(0, 2, 3)
x0 <- fit.full$Xp
for(j in 1:3) {
  re <- fit.full$y - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  lim.rob[,j] <- c(min(re), max(re))
  plot(re ~ x0[,j], type='p', pch=19, col='gray30', 
       xlab=colnames(x0)[j], ylab='', cex=1.5)
  oo <- order(x0[,j])
  lines(x0[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

NOTE: These plots could also be obtained using the `plot` method: 
```{r plot.method, eval=FALSE}
# Plot each component
plot(fit.full, which=1:3)
```

<!---
Aca iria cv para gam

--->

The optimal bandwidths are `0.7`, `0.7` and `0.5` for `Solar.R`, 
`Wind` and `Temp`, respectively. The code to compute the leave-one-out
cross-validation is available in the README file in the GitHub repository.
Below are plots of  partial residuals with
the classical and robust fits overlaid:
```{r classicfits, fig.show="hold", out.width="33%"}
aircomplete <- airquality[ complete.cases(airquality), ]
library(gam)
fit.gam <- gam(Ozone ~ lo(Solar.R, span=.7) + lo(Wind, span=.7)+
                 lo(Temp, span=.5), data=aircomplete)
# Plot both fits (robust and classical) 
x <- as.matrix( aircomplete[ , c('Solar.R', 'Wind', 'Temp')] )
y <- as.vector( aircomplete[ , 'Ozone'] )
fits <- predict(fit.gam, type='terms')
# alpha.gam <- attr(fits, 'constant')
for(j in 1:3) {
  re <- fit.full$yp - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=20, col='gray45', xlab=colnames(x)[j], ylab='')
  oo <- order(x[,j])
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
  lines(x[oo,j], fits[oo,j], lwd=5, col='magenta')
}
```

To identify potential outiers 
we look at the residuals from the robust fit, and use
the function `boxplot`:
```{r outliers, fig.width=4, fig.height=4}
re.ro <- residuals(fit.full)
ou.ro <- boxplot(re.ro, col='gray80')$out
in.ro <- (1:length(re.ro))[ re.ro %in% ou.ro ]
points(rep(1, length(in.ro)), re.ro[in.ro], pch=20, cex=1.5, col='red')
```

We highlight these suspicious observations on the
scatter plot 
```{r showouts}
cs <- rep('gray30', nrow(aircomplete))
cs[in.ro] <- 'red'
os <- 1:nrow(aircomplete)
os2 <- c(os[-in.ro], os[in.ro])
pairs(aircomplete[os2, c('Ozone', 'Solar.R', 'Wind', 'Temp')], 
      pch=19, col=cs[os2], cex=1.5)
```

and on the partial residuals plots
```{r showouts2, fig.show="hold", out.width="33%"}
for(j in 1:3) {
  re <- fit.full$yp - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=20, col='gray45', xlab=colnames(x)[j], ylab='')
  points(re[in.ro] ~ x0[in.ro,j], pch=19, col='red', cex=1.5)
  oo <- order(x[,j])
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
  lines(x[oo,j], fits[oo,j], lwd=5, col='magenta')
}
```

We now compute the classical backfitting algorithm on
the data without the potential outliers identified by
the robust fit (the optimal smoothing 
parameters for the non-robust fit were 
re-computed using leave-one-out cross-validation on the "clean" data set). 
Note that now both fits (robust and non-robust) are  
almost identical.
```{r bothonclean, fig.show="hold", out.width="33%"}
# Run the classical backfitting algorithm without outliers
airclean <- aircomplete[-in.ro, ]
fit.gam2 <- gam(Ozone ~ lo(Solar.R, span=.7) + lo(Wind, span=.8)+
                  lo(Temp, span=.3), data=airclean)
fits2 <- predict(fit.gam2, type='terms')
# alpha.gam2 <- attr(fits2, 'constant')
dd2 <- aircomplete[-in.ro, c('Solar.R', 'Wind', 'Temp')]
for(j in 1:3) {
  re <- fit.full$yp - fit.full$alpha - rowSums(fit.full$g.matrix[,-j])
  plot(re ~ x[,j], type='p', pch=20, col='gray45', xlab=colnames(x)[j], ylab='')
  points(re[in.ro] ~ x[in.ro,j], pch=20, col='red', cex=1.5)
  oo <- order(dd2[,j])
  lines(dd2[oo,j], fits2[oo,j], lwd=5, col='magenta')
  oo <- order(x[,j])
  lines(x[oo,j], fit.full$g.matrix[oo,j], lwd=5, col='blue')
}
```

<!--- Y aca desde que dice Finally

--->